### 2022-03-03 ehoh/ehoh.tex
background
jasmin unif

\blfootnote{In this work I desgined, implemented and evaluated all changes to term representation, algorithms and indexing data structures.}

@@

\begin{abstract}
Decades of work have gone into developing efficient proof calculi, data
structures, algorithms, and heuristics for first-order automatic theorem
proving. Higher-order provers lag behind in terms of efficiency. Instead of
developing a new higher-order prover from the ground up, we propose to start
with the state-of-the-art superposition prover E and gradually enrich it with
higher-order features. We explain how to extend the prover's data structures,
algorithms, and heuristics to $\lambda$-free higher-order logic, a formalism
that supports partial application and applied variables. Our extension
outperforms the traditional encoding and appears promising as a stepping stone
toward full higher-order logic.    
\end{abstract}

@@

\looseness=-1
A central aspect of our work is a set of techniques we call
\emph{prefix optimization}. Taking a traditional look at higher-order terms, they contain twice as many proper
subterms as first-order terms; for example,
$\cst{f}\;(\cst{g}\;\cst{a})\;\cst{b}$ contains not only the ``argument'' subterms
$\cst{g}\;\cst{a}$, $\cst{a}$, $\cst{b}$ but also the ``prefix'' subterms
$\cst{f}$, $\cst{f}\;(\cst{g}\;\cst{a})$, $\cst{g}$.
\begin{rep}Many operations, including superposition and rewriting, require
traversing all subterms of a term.\end{rep}
Using the optimization, the prover traverses subterms recursively in a
first-order fashion, considering all the prefixes of a given subterm
together. %, at little additional cost.
%
Our experiments (Sect.~\ref{sec:ehoh:evaluation}) show that Ehoh is
almost as fast as E on first-order problems and can also prove
higher-order problems that do not require synthesizing
$\lambda$-terms. As next steps, we plan to add support for
$\lambda$-terms and higher-order unification.

@@

\looseness=-1
Our logic is a variant of the intensional $\lambda$-free Boolean-free higher-order
logic (\lfhol{}) described by Bentkamp et al.\ %, Blanchette, Cruanes, and Waldmann
\cite[Sect.~2]{bbcw-21-lfho}, which could also be called ``applicative
first-order logic.'' In the spirit of FOOL \cite{kotelnikov-16-fool}, we
%% kotelnikov-15-fool
extend the syntax of this logic by erasing the distinction between terms and
formulas, and its semantics by interpreting the Boolean type $o$ as a domain
of cardinality~2. Functional extensionality can be obtained by adding
suitable axioms \cite[Sect.~3.1]{bbcw-21-lfho}.

@@

This logic differs from the higher-order logic described in
Sect.~\ref{sec:pre:hol} three ways. First, $\lambda$-abstraction is disallowed.
Second, logical connectives are not part of the set of symbols. Instead, there
is a special inductive case in the definition of terms that defines formulas.
Third, subterms are defined in a more traditional way.

\looseness=-1
For completeness, we provide the definition of terms. Terms, ranged over by $s,t,u,v$, are either
\emph{variables} $x, y, z, \dots$, (\emph{function}) \emph{symbols} $\cst{a},
\cst{b}, \cst{c}, \cst{d},\allowbreak \cst{f},\allowbreak \cst{g}, \ldots{}$
(often called ``constants'' in the higher-order literature), binary applications
$s \; t$, or Boolean terms $\itrue$, $\ifalse$, $\inot s$, $s \mathbin{\iand}
t$, $s \mathbin{\ior} t$, $s \mathbin{\iimplies} t$, $s \mathbin{\iequiv} t$,
$\iforall x.\> s$, $\iexists x.\> s$, $s \mathrel{\ieq} t$. E and Ehoh clausify
the input as a preprocessing step, producing a clause set in which the only
proper Boolean subterms are variables, $\itrue$, and $\ifalse$. Note that we use
lowercase letters for free variables as bound variables do not appear in
clauses. A term's \emph{arity} is the number of extra arguments it can take. If
$\iota$ is a base type, $\cst{f}$ has type $\iota \to \iota \to \iota$, and
$\cst{a}$ has type $\iota$, then $\cst{f}$ is binary, $\cst{f}\;\cst{a}$ is
unary, and $\cst{f}\;\cst{a}\;\cst{a}$ is nullary. Subterms are defined in the
traditional higher-order way; for example, $s \; t$ has all subterms of $s$ and
$t$ as subterms, in addition to $s \; t$ itself; as a consequence
$\cst{f} \; \cst{a}$ is a subterm of $\cst{f} \; \cst{a} \;\cst{a}$. With $\Var(x)$
we denote the set of free variables of $x$ where $x$ ranges over terms, clauses,
set of clauses or any other objects that contain terms.

@@

\begin{rep}The TPTP includes benchmarks from various areas of
computer science and mathematics. It is the de facto standard for evaluating
automatic provers, but it has few higher-order problems.
%
\end{rep}%
For the first group of benchmarks, we randomly selected 1000 FOF problems
(out of 8172)
and all monomorphic TFF problems that are parsable by E\confrep{}{ within 60\,s
(amounting to 147 out of 231 monomorphic TFF problems)}.
%
Both groups of Sledgehammer problems include two subgroups of 2506 problems,
generated to include 32 or 512 Isabelle lemmas (SH32 and SH512), to
represent both small and large problems.
% arising in interactive
%verification.
Each subgroup consists of two sub-sub\-groups of 1253
problems, generated by using either $\lambda$-lifting or SK-style combinators
to encode $\lambda$-expressions.

\looseness=-1
To ascertain the effectiveness of our approach, we evaluated Ehoh against
E used on applicative encodings of problems (denoted by \appE{}). For reference,
we also evaluated the following versions of higher-order provers that competed in the
THF division of the 2019 edition of CASC \cite{gs-19-casc27}:\
CVC4
1.8 prerelease \cite{brotb-19-ho-smt}, Leo-III 1.4 \cite{sb-21-leo3},
Satallax 3.4 \cite{cb-12-satallax}, Vampire 4.4 \cite{br-19-restricted-unif},
and Zipperposition 1.6 \cite{bbtvw-21-sup-lam}.
Like at CASC, we used different versions of Vampire for first-order
and higher-order problems. Similarly, Zipperposition does not use E as backend when
it is run on first-order problems
and uses different heuristics on first-~and higher-order problems.
The genuine higher-order provers have
the unfair advantage that they can instantiate higher-order
variables with $\lambda$-terms. % (or equivalent SK-style combinator terms).
Thus, some formulas that are
provable by these systems may be nontheorems for \appE{} and Ehoh\confrep{}{, or
they may require tedious reasoning about $\lambda$-lifted functions or SK-style
combinators}.
An example is the conjecture $\exists f.\>\forall x\> y.\; f\>x\;y \eq
\cst{g}\;y\;x$, whose proof requires taking $\lambda x\> y.\; \cst{g}\;y\;x$
as the witness for~$f$.

@@

We observe the following.
%
First, comparing the Ehoh row with the E row, we see that Ehoh's
  overhead is barely noticeable---the difference is at most \NumberOK{two}
  problems.
  %The raw evaluation data reveal that Ehoh's time overhead is about \NumberNOK{3.7\%}.
%
Second, Ehoh outperforms the applicative encoding on both first-order
  and higher-order problems. Nevertheless, the raw evaluation data reveal that
  there are quite a few higher-order problems that \appE{}
  proves faster than Ehoh. %\confrep{}{, because the applicative encoding impacts the
%  heuristics in subtle ways}.
%
Third, it is advantageous to use the higher-order versions of the
  Sledgehammer problems, although the difference in success rate is
  small, especially for SH512.
%
Fourth, the new \emph{hoboa} %configuration
  outperforms \emph{boa} on
  higher-order problems, suggesting that it could be worthwhile to re-train
  \emph{auto} and \emph{autoschedule} based on \lfhol{} benchmarks
  and to design further heuristics.
%
Fifth, Ehoh cannot compete against the best higher-order systems, but this
  is no surprise given that it does not yet support $\lambda$-expressions and
  higher-order unification.

@@

\begin{rep}
  Our working hypothesis is that it is possible to extend %existing
  first-order provers to higher-order logic without slowing them down unduly.
  Our research program is two-pronged:
  On the theoretical side, we are
  investigating higher-order extensions of superposition
  \cite{bbcw-21-lfho,bbtvw-21-sup-lam,vn-20-bools};
  on the practical side, we are implementing such extensions in a
  state-of-the-art prover.
  %TODO_PHD: another reference

  The work described in this chapter required modifying many parts of the E
  prover. The invariant that variables cannot be applied and that symbols are always
  passed the same number of arguments were entrenched in E's code,
  requiring hundreds of modifications. Nonetheless,
  %with a good understanding of the prover's inner working,
  %% Obvious, since Stephan is a coauthor. --JB
  we found the generalization manageable and are now in a position to add
  support for $\lambda$-terms and higher-order unification.
  \end{rep}

  \begin{sloppypar}
  Traditionally, most higher-order provers were designed from the ground up
  to target higher-order logic. Two
  exceptions are Otter-$\lambda$ by Beeson \cite{mb-04-lam-logic} and
  Zipperposition by Cruanes et al.\ %, Vukmirovi\'c, and colleagues
  \cite{sc-15-simon-phd,bbtvw-21-sup-lam}. \hbox{Otter-$\lambda$}
  adds $\lambda$-terms and second-order unification to
  the superposition prover Otter\begin{rep} \cite{mcc-03-otter}\end{rep}.
  % The approach is pragmatic. , with little emphasis on completeness.
  Zipperposition, also based on superposition, was extended
  to Bool\-e\-an-free higher-order logic by Bentkamp et al.\
  \cite{bbtvw-21-sup-lam}. Its performance is a far cry from E's, but
  it is easier to modify.
  Vukmirovi\'c et al.\ also used it to test and evaluate higher-order unification
  procedures \cite{vbn-21-unif} and Boolean reasoning
  \cite{vn-20-bools}.
  Zipperposition now includes Ehoh as a backend in a cooperative
  architecture.
  Finally, there is recent work by the developers of Vampire
  % br-20-full-sup-w-combs
  \cite{br-19-restricted-unif} and of the SMT
  (satisfiability modulo theories) solvers CVC4 and veriT
  \cite{brotb-19-ho-smt} to extend their provers to higher-order logic.
  \end{sloppypar}

@@

Despite considerable progress since the 1970s, higher-order automated
reasoning has not yet assimilated some of the most successful methods for
first-order logic with equality, such as superposition. We presented a
graceful extension of a state-of-the-art first-order theorem prover to a
fragment of higher-order logic devoid of $\lambda$-terms. Our work
covers both theoretical and practical aspects. Experiments show promising
results on $\lambda$-free higher-order problems and very little overhead for
first-order problems, as we would expect from a graceful generalization.

>>>>

\blfootnote{In this work I designed, implemented and evaluated all changes to
term representation, algorithms and indexing data structures. Jasmin Blanchette
came up with the extension of fingerprinting indexing. Stephan Schulz provided
the necessary E expertise.}

@@

\begin{abstract}
Decades of work have gone into developing efficient proof calculi, data
structures, algorithms, and heuristics for first-order automatic theorem
proving. Higher-order provers lag behind in terms of efficiency. Instead of
developing a new higher-order prover from the ground up, we propose to start
with the state-of-the-art superposition prover E and gradually enrich it with
higher-order features. We explain how to extend the prover's data structures,
algorithms, and heuristics to $\lambda$-free higher-order logic, a formalism
that supports partial application and applied variables. Our extension
outperforms the traditional encoding and forms a stepping stone
toward full higher-order logic.    
\end{abstract}

@@

\looseness=-1
A central aspect of our work is a set of techniques we call
\emph{prefix optimization}. Taking a traditional look at higher-order terms, they contain twice as many proper
subterms as first-order terms; for example,
$\cst{f}\;(\cst{g}\;\cst{a})\;\cst{b}$ contains not only the ``argument'' subterms
$\cst{g}\;\cst{a}$, $\cst{a}$, $\cst{b}$ but also the ``prefix'' subterms
$\cst{f}$, $\cst{f}\;(\cst{g}\;\cst{a})$, $\cst{g}$.
\begin{rep}Many operations, including superposition and rewriting, require
traversing all subterms of a term.\end{rep}
Using the optimization, the prover traverses subterms recursively in a
first-order fashion, considering all the prefixes of a given subterm
together. %, at little additional cost.
%
Our experiments (Sect.~\ref{sec:ehoh:evaluation}) show that Ehoh is
almost as fast as E on first-order problems and can also prove
higher-order problems that do not require synthesizing
$\lambda$-terms. In Chapter~\ref{ch:ehoh2} we show how to add support for
$\lambda$-terms and higher-order unification.

@@

\looseness=-1
Our logic is a variant of the intensional $\lambda$-free Boolean-free
higher-order logic (\lfhol{}) described by Bentkamp et al.\
%, Blanchette, Cruanes, and Waldmann
\cite[Sect.~2]{bbcw-21-lfho}, which could also be called ``applicative
first-order logic.'' In the spirit of FOOL \cite{kotelnikov-16-fool}, we
%% kotelnikov-15-fool
extend the syntax of this by making formulas a special case of terms,
and its semantics by interpreting the Boolean type $o$ as a domain of
cardinality~2. Functional extensionality can be obtained by adding suitable
axioms \cite[Sect.~3.1]{bbcw-21-lfho}.

@@

This logic differs from the higher-order logic described in
Sect.~\ref{sec:pre:hol} in three ways. First, $\lambda$-abstraction is disallowed.
Second, logical connectives are not part of the set of symbols. Instead, there
is a special inductive case in the definition of terms that defines formulas.
Third, subterms are defined in a more traditional way, as defined below.

\looseness=-1
For reference, we provide the definition of terms. Terms, ranged over by $s,t,u,v$, are either
\emph{variables} $x, y, z, \dots$, (\emph{function}) \emph{symbols} $\cst{a},
\cst{b}, \cst{c}, \cst{d},\allowbreak \cst{f},\allowbreak \cst{g}, \ldots{}$
(often called ``constants'' in the higher-order literature), binary applications
$s \; t$, or Boolean terms $\itrue$, $\ifalse$, $\inot s$, $s \mathbin{\iand}
t$, $s \mathbin{\ior} t$, $s \mathbin{\iimplies} t$, $s \mathbin{\iequiv} t$,
$\iforall x.\> s$, $\iexists x.\> s$, $s \mathrel{\ieq} t$. E and Ehoh clausify
the input as a preprocessing step, producing a clause set in which the only
proper Boolean subterms are variables, $\itrue$, and $\ifalse$. Note that we use
lowercase letters for free variables as bound variables do not appear in
clauses. A term's \emph{arity} is the number of extra arguments it can take. If
$\iota$ is a base type, $\cst{f}$ has type $\iota \to \iota \to \iota$, and
$\cst{a}$ has type $\iota$, then $\cst{f}$ is binary, $\cst{f}\;\cst{a}$ is
unary, and $\cst{f}\;\cst{a}\;\cst{a}$ is nullary. Subterms are defined in the
traditional higher-order way; for example, $s \; t$ has all subterms of $s$ and
$t$ as subterms, in addition to $s \; t$ itself; as a consequence
$\cst{f} \; \cst{a}$ is a subterm of $\cst{f} \; \cst{a} \;\cst{a}$. With $\Var(x)$
we denote the set of free variables of $x$ where $x$ ranges over terms, clauses,
set of clauses, or any other objects that contain terms.

@@

\begin{rep}The TPTP includes benchmarks from various areas of
computer science and mathematics. It is the de facto standard for evaluating
automatic provers, but it has few higher-order problems.
%
\end{rep}%
For the first group of benchmarks, we randomly selected 1000 FOF problems
(out of 8172)
and all monomorphic TFF problems that are parsable by E\confrep{}{ within 60\,s
(amounting to 147 out of 231 monomorphic TFF problems)}.
%
Both groups of Sledgehammer problems include two subgroups of 2506 problems,
generated to include 32 or 512 Isabelle lemmas (SH32 and SH512), to
represent both small and large problems.
% arising in interactive
%verification.
Each subgroup consists of two sub-sub\-groups of 1253
problems, generated by using either $\lambda$-lifting or SKBCI-style combinators
to encode $\lambda$-expressions.

\looseness=-1
To ascertain the effectiveness of our approach, we evaluated Ehoh against
E used on applicative encodings of problems (denoted by \appE{}). For reference,
we also evaluated the following versions of higher-order provers that competed in the
THF division of the 2019 edition of CASC \cite{gs-19-casc27}:\
CVC4
1.8 prerelease \cite{brotb-19-ho-smt}, Leo-III 1.4 \cite{sb-21-leo3},
Satallax 3.4 \cite{cb-12-satallax}, Vampire 4.4 \cite{br-19-restricted-unif},
and Zipperposition 1.6 \cite{bbtvw-21-sup-lam}.
Like at CASC, we used different versions of Vampire for first-order
and higher-order problems. Similarly, Zipperposition does not use E as backend when
it is run on first-order problems
and uses different heuristics on first-~and higher-order problems.
The genuine higher-order provers have
the unfair advantage that they can instantiate higher-order
variables with $\lambda$-terms. % (or equivalent SK-style combinator terms).
Thus, some formulas that are
provable by these systems may be nontheorems for \appE{} and Ehoh\confrep{}{, or
they may require tedious reasoning about $\lambda$-lifted functions or SKBCI-style
combinators}.
An example is the conjecture $\exists f.\>\forall x\> y.\; f\>x\;y \eq
\cst{g}\;y\;x$, whose proof requires taking $\lambda x\> y.\; \cst{g}\;y\;x$
as the witness for~$f$.

@@

We observe the following.
%
First, comparing the Ehoh row with the E row, we see that Ehoh's
  overhead is barely noticeable---the difference is at most \NumberOK{two}
  problems.
  %The raw evaluation data reveal that Ehoh's time overhead is about \NumberNOK{3.7\%}.
%
Second, Ehoh outperforms the applicative encoding on both first-order
  and higher-order problems. Nevertheless, the raw evaluation data reveal that
  there are quite a few higher-order problems that \appE{}
  proves faster than Ehoh. %\confrep{}{, because the applicative encoding impacts the
%  heuristics in subtle ways}.
%
Third, it is advantageous to use the higher-order versions of the
  Sledgehammer problems, although the difference in success rate is
  small, especially for SH512.
%
Fourth, the new \emph{hoboa} %configuration
  outperforms \emph{boa} on
  higher-order problems, suggesting that it could be worthwhile to re-train
  \emph{auto} and \emph{autoschedule} based on \lfhol{} benchmarks
  and to design further heuristics.
%
Fifth, Ehoh cannot compete against the best higher-order systems, but this
  is no surprise given that it does not support $\lambda$-expressions and
  higher-order unification. An extension of Ehoh that supports these features
  is described in Chapter~\ref{ch:ehoh2}.

@@

\begin{rep}
  Our working hypothesis is that it is possible to extend %existing
  first-order provers to higher-order logic without slowing them down unduly.
  Our research program is two-pronged:
  On the theoretical side, we are
  investigating higher-order extensions of superposition
  \cite{bbcw-21-lfho,bbtvw-21-sup-lam,vn-20-bools};
  on the practical side, we are implementing such extensions in a
  state-of-the-art prover. In this thesis, the focus is on the second aspect.
  %TODO_PHD: another reference

  The work described in this chapter required modifying many parts of the E
  prover. The invariant that variables cannot be applied and that symbols are always
  passed the same number of arguments were entrenched in E's code,
  requiring hundreds of modifications. Nonetheless,
  %with a good understanding of the prover's inner working,
  %% Obvious, since Stephan is a coauthor. --JB
  we found the generalization manageable and are now in a position to add
  support for $\lambda$-terms and higher-order unification, as discussed in Chapter~\ref{ch:ehoh2}.
  \end{rep}

  \begin{sloppypar}
  Traditionally, most higher-order provers were designed from the ground up
  to target higher-order logic. Two
  exceptions are Otter-$\lambda$ by Beeson \cite{mb-04-lam-logic} and
  Zipperposition by Cruanes et al.\ %, Vukmirovi\'c, and colleagues
  \cite{sc-15-simon-phd,bbtvw-21-sup-lam}. \hbox{Otter-$\lambda$}
  adds $\lambda$-terms and second-order unification to
  the superposition prover Otter\begin{rep} \cite{mcc-03-otter}\end{rep}.
  % The approach is pragmatic. , with little emphasis on completeness.
  Zipperposition, also based on superposition, was extended
  to Bool\-e\-an-free higher-order logic by Bentkamp et al.\
  \cite{bbtvw-21-sup-lam}. Its performance is a far cry from E's, but
  it is easier to modify.
  Vukmirovi\'c et al.\ also used it to test and evaluate higher-order unification
  procedures \cite{vbn-21-unif} and Boolean reasoning
  \cite{vn-20-bools}.
  Zipperposition now includes Ehoh as a backend in a cooperative
  architecture.
  Finally, there is work by the developers of Vampire
  % br-20-full-sup-w-combs
  \cite{br-19-restricted-unif} and of the SMT
  (satisfiability modulo theories) solvers CVC4 and veriT
  \cite{brotb-19-ho-smt} to extend their provers to higher-order logic.
  \end{sloppypar}

@@

Despite considerable progress since the 1970s, until a few years ago higher-order automated
reasoning has not assimilated some of the most successful methods for
first-order logic with equality, such as superposition. We presented a
graceful extension of a state-of-the-art first-order theorem prover to a
fragment of higher-order logic devoid of $\lambda$-terms. Our work
covers both theoretical and practical aspects. Experiments show promising
results on $\lambda$-free higher-order problems and very little overhead for
first-order problems, as we would expect from a graceful generalization.
